Supporting information: Copulas and their potential for ecology

### 2019-10-07 SupportingInformation.Rmd
finished an edit of Discussion, made a list of what else needs to be done

# Hypothetical evolutionary mechanism for tail dependence between traits \label{evol_mech}

@@

<!--***DAN: Move all the rest of it into the Discussion, I think, but can decide when I get to the Discussion. If it is mved, then you may want to retitle this appendix to reflect the 
fact it is onl methods-->
Performing statistical tests for associations between continuous traits in different species
    was a primary motivating example of phylogenetic comparative methods.
For example, Felsenstein's method of phylogenetically independent contrasts 
    [@Felsenstein1985] is currently the second most-cited paper in the history
    of the journal the American Naturalist [@HueyGT2019].
Frustratingly, the field still lacks a dependable procedure for dealing with branch lengths, 
    which are a crucial input to the method.
Ideally, the branch lengths used to correct for phylogenetic effects would represent
    the expected amounts of change for the characters that are being analyzed.
Because researchers almost never have a reliable method for providing such branch lengths,
    most researchers rely on ultrametric trees -- those for which the branch
    length can be treated as a proxy for the duration of the branch in time.
Frequently these branch lengths are transformed to assess sensitivity to different 
    assumptions about the degree of phylogenetic inertia displayed by the traits
    under study [@Ives2018;@Harmon2018].
Even if one were able to simply use a time-based set of branch lengths, assigning
    dates to nodes in phylogenies is difficult.
DNA sequence data can provide estimates of branch lengths, but these estimates
    are dependent on the adequacy of models which correct sequences for multiple
    substitutions occurring at the same location.
Biases in estimating the evolutionary distance can affect downstream analyses [@Phillips2009].
Additionally, changes
    in the rate of molecular evolution make the estimation of dates difficult [@HeathM2014] even when branch lengths in time are accurately estimated.

Without reliable branch length estimates, it is difficult to interpret the significance
    of the magnitude of changes in traits across a tree.
Developing tests of association based on copula structure may possibly lead the way to more
    robust methods for studying associations when we lack defensible estimates of
    branch lengths. 
We note that our phylogenetic analyses here consisted merely of simulations to assess
    whether interesting copula structure in a simple evolutionary process could 
    leave a detectable signal on the trait data for extant species.
Substantial work remains to be done before we have a copula-based method for 
    analyzing data on a phylogenetic tree. 
But the nonparametric nature of approaches
    based solely on data ranks, as some of our copula methods are, may be a promising avenue for 
    avoiding inaccuracies that arise via the highly structured model assumptions implicit
    in the method of phylogenetically independent contrasts and related methods.

Additionally, character evolution was simulated using one random draw from the relevant matrix
per phylogeny branch. This was principally because branch lengths are often hugely
uncertain. If branch lengths were well known, alternative simulation strategies may
include selecting one draw from the matrix of evolutionary changes per unit time, 
or selecting one draw for the whole branch but rescaling the variances of the selected
character changes according to the 
length of the branch. These choices amount to the same thing for normal copula
structure, but not otherwise. Modelling choices such as these may have influenced
our results. Additional research also seems warranted testing the realism 
of our hypothesized mechanism and simulations.

@@

We examined using simulations whether copula structure influences spatial Taylor's law, as 
follows. First, an $n \times n$ covariance matrix $\Sigma$ was created
with unit diagonal entries 
and off-diagonal entries $\rho=0.7$; $n=25$ was the number of locations at which populations
were sampled. Then $N \times 1000$ random vectors were generated from a multivariate normal 
distribution with mean vector $(0,\ldots,0)$ and covariance matrix $\Sigma$. These 
were organized into an $N \times n \times 1000$ array representing $1000$ replicate 
simulations of (temporally independent, in this example) 
spatio-temporal population dynamics measured in 
$n$ locations over $N$ time steps. We used $N=50$.
The cumulative distribution function
(cdf) of the standard normal distribution was applied to these numbers, followed by the 
inverse of the cdf of a gamma distribution with shape and scale parameters $7.5$ and $1$. 
These transformations made gamma distributed the marginal distributions of populations in 
each sampling location for each simulation. 
A skewed population marginal distribution is an 
important factor in Taylor's law [@Cohen2015; @Reuman2017]. We refer henceforth to the 
resulting $N \times n \times 1000$ array as our *normal-copula population array* 
for this simulation exercise;
copula structure of dependences between populations in different locations
was normal for this array.

We then generated a series of arrays that were the same dimensions ($N \times n \times 1000$)
as the normal-copula population array, and that were statistically similar to 
the normal-copula population array in all ways except the new
arrays were constructed to have Clayton (respectively, Frank, survival Clayton) copula structure for 
the dependences between 
populations in different sampling locations within a simulation. We 
refer to the new arrays as our
*Clayton-copula*, *Frank-copula*, and *Survival Clayton-copula population arrays*, respectively. The Clayton-copula array was
constructed as follows; the others were constructed analogously. First, we constructed an
$n$-dimensional Clayton copula with parameter chosen so that the 
Spearman correlations between each pair of the $n$ components were the same as those
of the normal distribution constructed above. This was done using the `rho` and `iRho` functions 
in the `copula` package in R. Because the Clayton copula is a one-paremeter family, this 
step relies on the fact that all pairs of locations were 
equally correlated with each other in the original multivariate normal distribution
constructed above. We then randomized each simulation in 
the normal-copula population array in a special way 
(see Appendix \ref{sect:copsurrognd}) so that the resulting 
surrogate/randomized dataset had the copula structure
of the Clayton copula just constructed. The randomization procedure worked 
by permuting time series of values in the normal-copula array, separately for each 
sampling location and simulation. So population marginal distrbutions for each location
and simulation were *exactly* the same as those in the normal-copula population array. 
Permutations were done in such a way
that resulting pairwise Spearman correlations were the same as the Clayton copula 
constructed above, and hence were the same as the normal-copula array, 
except for sampling variation. Thus, to summarize using mathematical notation,
if $m^{(\text{N})}$ is our normal-copula population array and $m^{(\text{C})}$ is our 
Clayton-copula population array, then,
for any given $i$ and $j$, the sets $\{m^{(\text{N})}_{tij}: t=1,\ldots,N\}$ and 
$\{m^{(\text{C})}_{tij}: t=1,\ldots,N\}$ are exactly the same (when considered 
as unordered sets); 
and for any given $i_1$, $i_2$ and $j$, the Spearman correlations $\cor_t(m^{(\text{C})}_{ti_1j},m^{(\text{C})}_{ti_2j})$
and $\cor_t(m^{(\text{N})}_{ti_1j},m^{(\text{N})}_{ti_2j})$ are the same, to within 
sampling variation. The whole exercise was repeated preserving Kendall instead of 
Spearman correlations.

A variety of Taylor's law statistics were computed for each simulation $m^{(X)}_{tij}$ 
($i=1,\ldots,n$, $t=1,\ldots,N$) for each copula structure $X$ used. First, the $N=50$ 
spatial means,
$m_t$, and variances, $v_t$, were computed, and $\log(v_t)$ vs. $\log(m_t)$ plots 
were considered. As a shorthand, denote $y=\log(v_t)$ and $x=\log(m_t)$ for a 
single simulation. Linearity of the $y$ vs. $x$ relationship was 
tested for each simulation by comparing the linear regression
through these $50$ points to a quadratic alternative via an $F$-test, 
producing a $p$-value result. If 
these $p$-values were uniformly distributed across the unit interval for the 1000 replicate 
simulations, it supported the linearity assumption of Taylor's law; whereas if 
they were clustered toward smaller values
the test tended to reject that assumption. We also tested, for each simulation, 
the assumption of Taylor's 
law that the $y$ vs. $x$ plot was homoskedastic: we regressed the absolute residuals of the 
linear regression of $y$ versus $x$ against the predictions of that regression. 
A significant $p$-value
result of this test indicates heteroskedasticity. For each simulation we also computed the
root mean squared error of $y$ vs. $x$ data from the linear regression
of $y$ versus $x$, as well as the intercept and slope of the linear regression.
Finally we recorded the quadratic coefficient of the regression of $y$ against
$x$ and $x^2$, and the mean curvature of the quadratic regression equation itself
across the values $x$. The distrubutions of all these statistics across 1000 replicate 
simulations were compared for the Clayton, normal, Frank, and survival Clayton simulations to determine if copula structure 
influences Taylor's law.

Taylor's law was strongly influenced, and was often even invalidated in its assumptions of 
linearity and homoskedasticity, by non-normal copula structure. For normal copula structure 
(i.e., for simulations in our normal-copula population array), $p$-values for the linearty
and homoskedasticity tests were roughly uniformly distributed across replicate 
simulations and Taylor's law appeared visually to be a reasonable approximation
of the $\log(v_t)$ versus $\log(m_t)$ relationship (Fig. \ref{fig_spTL}B,E,F). 
Furthermore, quadratic coefficients
and curvature values were close to $0$ (Fig. \ref{fig_spTL}J,K), and root mean squared
errors from the linear regression were relatively small (Fig. \ref{fig_spTL}G).
But linearity or homoskedasticity were violated more frequently for non-normal copula structure
(Fig. \ref{fig_spTL}A,C,D,E,F); quadratic coefficients and curvatures were frequently 
non-zero (Fig. \ref{fig_spTL}J,K); and root mean squared
errors from the linear regression were much higher (Fig. \ref{fig_spTL}G). Slopes 
and intercepts of the linear regression were also strongly affected by copula
structure (Fig. \ref{fig_spTL}H,I), though some of the effect here was because
linear regressions do not always adequately represent the $\log(v_t)$ versus $\log(m_t)$
relationship when copula structure was not normal.

@@

# From causes to consequences for green spruce aphid abundance, details \label{sect:greenspruce}
Winter temperature for year $t$ was for December of year $t-1$ to March of 
year $t$. Temperature data, which pertained to the same locations and years as the aphid data,
were obtained and processed exactly as in @Sheppard2016. Copulas
describing the relationship between green spruce aphid abundance and winter 
temperature are in Fig. \ref{fig:greenspruce}. Results of analyses of these
copulas that paralleled the analyses applied to bivariate ecological datasets 
(see section on methods for Q1) are in Table \ref{tab:greenspruce}.

@@

res<-ms_npa_RES_greenspruceaphid_count_temp$sp10_Green_spruce_aphid$ms_summary_allloc
#row.names(res)<-NULL
#res<-cbind(c(2:11),res)
colnames(res)<-c("p-value, independence test","Best-fit copula",
                 "Best-fit copula AICw","Normal copula AICw","p, CvM","p, KS","Model-avg. LT","Model-avg. UT","Model-avg. LT minus UT","cor$_{l}$ - cor$_{u}$","P$_{l}$ - P$_{u}$","D$_{u}^2$ - D$_{l}^2$")
res<-res[,-1]
knitr::kable(res,"latex", booktabs = T, linesep = "",digits=3, escape=F,longtable=T,
             caption.short="Summary results for copulas for winter temperature versus green spruce aphid abundance",
      caption = "Summary results for analysis of copulas for winter temperature versus green spruce 
aphid abundance for each of the 10 sampling locations. The column labeled p, CvM is the p-value 
result for the goodness of fit test using the Cramer-von Mises statistic (see section on 
methods for Q1). The 
column labeled p, KS is the p-value 
result for the goodness of fit test using the Kolmogorov-Smirnov statistic. Entries with an NA are because model selection and other statistics
were only employed for copulas for which independence was rejected (5 percent significance level).\\label{tab:greenspruce}") %>% column_spec(1:12,width="1.2 cm")#%>%
#kable_styling(latex_options = c("scale_down"))
```

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm]{./Results/aphid_count_winter_temp_results/copulaplot_greenspruceaphidcount_temp.pdf}
\caption[Copula plots for winter temperature versus green spruce aphid abundance]{Copula plots 
for winter temperature versus green spruce 
aphid abundance for each of the 10 sampling locations. The $p$-values in the header are from the test of independence (see section on methods for Q1).\label{fig:greenspruce}}
\end{center}
\end{figure}

>>>>

# Methods for the hypothetical evolutionary mechanism for tail dependence between traits \label{evol_mech}

@@

@@

As described in the main text, we examined using simulations 
whether copula structure influences spatial Taylor's law; details are provided here. 
First, an $n \times n$ covariance matrix $\Sigma$ was created
with unit diagonal entries 
and off-diagonal entries $\rho=0.7$; $n=25$ was the number of locations at which populations
were sampled. Then $N \times 1000$ random vectors were generated from a multivariate normal 
distribution with mean vector $(0,\ldots,0)$ and covariance matrix $\Sigma$. These 
were organized into an $N \times n \times 1000$ array representing $1000$ replicate 
simulations of (temporally independent, in this example) 
spatio-temporal population dynamics measured in 
$n$ locations over $N$ time steps. We used $N=50$.
The cdf of the standard normal distribution was applied to these numbers, followed by the 
inverse of the cdf of a gamma distribution with shape and scale parameters $7.5$ and $1$. 
These transformations made gamma distributed the marginal distributions of populations in 
each sampling location for each simulation. 
A skewed population marginal distribution is an 
important factor in Taylor's law [@Cohen2015; @Reuman2017]. We refer henceforth to the 
resulting $N \times n \times 1000$ array as our *normal-copula population array* 
for this simulation exercise;
copula structure of dependences between populations in different locations
was normal for this array. The array contains $1000$ population matrices.

We then generated a series of arrays that were the same dimensions ($N \times n \times 1000$)
as the normal-copula population array, and that were statistically similar to 
the normal-copula population array in all ways except the new
arrays were constructed to have Clayton (respectively, Frank, survival Clayton) copula structure for 
the dependences between 
populations in different sampling locations within a simulation. We 
refer to the new arrays as our
*Clayton-copula*, *Frank-copula*, and *Survival Clayton-copula population arrays*, respectively. The Clayton-copula array was
constructed as follows; the others were constructed analogously. First, we constructed an
$n$-dimensional Clayton copula with parameter chosen so that the 
Spearman correlations between each pair of the $n$ components were the same as those
of the normal distribution constructed above. This was done using the `rho` and `iRho` functions 
in the `copula` package in R. Because the Clayton copula is a one-paremeter family, this 
step relies on the fact that all pairs of locations were 
equally correlated with each other in the original multivariate normal distribution
constructed above. We then randomized each simulation in 
the normal-copula population array in a special way 
(see Appendix \ref{sect:copsurrognd}) so that the resulting 
surrogate/randomized dataset had the copula structure
of the Clayton copula just constructed. The randomization procedure worked 
by permuting time series of values in the normal-copula array, separately for each 
sampling location and simulation. So population marginal distrbutions for each location
and simulation were *exactly* the same as those in the normal-copula population array. 
Permutations were done in such a way
that resulting pairwise Spearman correlations were the same as the Clayton copula 
constructed above, and hence were the same as the normal-copula array, 
except for sampling variation. Thus, to summarize using mathematical notation,
if $m^{(\text{N},k)}_{t,j}$ is our normal-copula population array (the three dimensions 
of the array are indexed, respectively, by $t$, $j$ and $k$, and $m^{(\text{C},k)}_{t,j}$ is our 
Clayton-copula population array, then,
for any given $j$ and $k$, the sets $\{m^{(\text{N},k)}_{t,j}: t=1,\ldots,N\}$ and 
$\{m^{(\text{C},k)}_{t,j}: t=1,\ldots,N\}$ are exactly the same (when considered 
as unordered sets); 
and for any given $j_1$, $j_2$ and $k$ with $j_1 \neq j_2$, the Spearman correlations $\cor_t(m^{(\text{C},k)}_{t,j_1},m^{(\text{C},k)}_{t,j_2})$
and $\cor_t(m^{(\text{N},k)}_{t,j_1},m^{(\text{N},k)}_{t,j_2})$ are the same, to within 
sampling variation. The whole exercise was repeated preserving Kendall instead of 
Spearman correlations.

@@

<!--Moved this stuff to main text for AER
# From causes to consequences for green spruce aphid abundance, details \label{sect:greenspruce}
Winter temperature for year $t$ was for December of year $t-1$ to March of 
year $t$. Temperature data, which pertained to the same locations and years as the aphid data,
were obtained and processed exactly as in @Sheppard2016. Copulas
describing the relationship between green spruce aphid abundance and winter 
temperature are in Fig. \ref{fig:greenspruce}. Results of analyses of these
copulas that paralleled the analyses applied to bivariate ecological datasets 
(see section on methods for Q1) are in Table \ref{tab:greenspruce}.
-->

@@

res<-ms_npa_RES_greenspruceaphid_count_temp$sp10_Green_spruce_aphid$ms_summary_allloc
#row.names(res)<-NULL
#res<-cbind(c(2:11),res)
colnames(res)<-c("p-value, independence test","Best-fit copula",
                 "Best-fit copula AICw","Normal copula AICw","p, CvM","p, KS","Model-avg. LT","Model-avg. UT","Model-avg. LT minus UT","cor$_{l}$ - cor$_{u}$","P$_{l}$ - P$_{u}$","D$_{u}^2$ - D$_{l}^2$")
res<-res[,-1]
knitr::kable(res,"latex", booktabs = T, linesep = "",digits=3, escape=F,longtable=T,
             caption.short="Summary results for copulas for winter temperature versus green spruce aphid abundance",
      caption = "Summary results for analysis of copulas for winter temperature versus green spruce 
aphid abundance for each of the 10 sampling locations, using the same model selection
and nonparametric methods detailed in the section on methods for Q1. The column labeled p, CvM is the p-value 
result for the goodness of fit test using the Cramer-von Mises statistic. The 
column labeled p, KS is the p-value 
result for the goodness of fit test using the Kolmogorov-Smirnov statistic. Entries with an NA are because model selection and other statistics
were only employed for copulas for which independence was rejected (5 percent significance level).\\label{tab:greenspruce}") %>% column_spec(1:12,width="1.2 cm")#%>%
#kable_styling(latex_options = c("scale_down"))
```

@@

<!--Decided to cut this for AER
\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm]{./Results/aphid_count_winter_temp_results/copulaplot_greenspruceaphidcount_temp.pdf}
\caption[Copula plots for winter temperature versus green spruce aphid abundance]{Copula plots 
for winter temperature versus green spruce 
aphid abundance for each of the 10 sampling locations. The $p$-values in the header are from the test of independence (see section on methods for Q1).\label{fig:greenspruce}}
\end{center}
\end{figure}
-->
